---
title: "PS1"
author: "Roshani Bulkunde"
date: "2024-01-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set working directory
```{r dir}
data_dir <-'C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Data'

setwd('C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Problem sets/PS1')
```

```{r}
# Load packages
library(tidyverse)
library(vtable)
library(readxl)
library(purrr)
```


## Wind turbine data
# Bring this data into R. You can use read.csv().
```{r}
WindTurbineR <- read_csv('C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Data/Wind Turbine/uswtdb_v6_0_20230531.csv')
```
##3. Remove observations with missing information on operation year, project capacity,
turbine capacity, turbine hub-height, turbine rotor diameter, longitude, and latitude
```{r}
#  To exclude observations that have missing data for only a subset of variables, create a logical vector that is a series of calls to is.na().
miss_wt <- is.na(WindTurbineR$p_year) |  
           is.na(WindTurbineR$p_cap) |  
           is.na(WindTurbineR$t_cap)  |  
           is.na(WindTurbineR$t_hh) |  
           is.na(WindTurbineR$t_rd) |  
           is.na(WindTurbineR$xlong) |  
           is.na(WindTurbineR$ylat)

# Count the number of rows flagged for deletion
sum(miss_wt)

# Use ! to include those that are NOT missing
WindTurbine1 <- subset(WindTurbineR,
                      subset = !miss_wt)

# Count the number of rows kept
nrow(WindTurbine1)
```
## 4. Remove observations for projects that started operating before 2001
```{r}
# WindTurbine2 <- subset(WindTurbine1, WindTurbine1$p_year > 2000)
WindTurbine2 <- WindTurbine1 %>% filter(p_year>2000)

# check the range of the year
range(WindTurbine2$p_year)
```
## 5.Remove observations from Alaska and Hawaii (we would only focus on contiguous
US).
```{r}
WindTurbineFinal <- WindTurbine2 %>% filter(!t_state %in% c("HI","AK"))

# convert kW to MW
WindTurbineFinal$t_cap <- WindTurbineFinal$t_cap/1000
```
##6. Create a summary table and show the mean, standard deviation, min, and max of
turbine capacity (in MW, not 1 MW = 1000 kW), turbine hub-height, turbine rotor
diameter. Also report the total number of turbines (i.e. N) in the table? *
```{r}
labs <- c('Turbine capacity (in MW)',
    'Turbine hub-height',
    'Turbine rotor diameter')

sumtable(WindTurbineFinal, vars = c('t_cap','t_hh', 't_rd'),
         summ=c('notNA(x)', 'mean(x)', 'sd(x)', 'min(x)', 'max(x)'),
         summ.names = list(c('N','Mean','SD','Min','Max')),
         labels=labs)



```
## 7. Aggregate/collapse the data to the project level. Use EIA ID for the aggregation.
#You can use several commands for aggregating data. The most common functions are: aggregate() or summarize() in package dplyr.
# While collapsing/aggregating the dataset to the project level, aggregate (or
keep depending on the variable) the following variables: mean project capacity
(in MW), operating year, total number of turbines in the project, mean
capacity of a turbine in the project (in MW), mean turbine hub-height, mean
turbine rotor diameter, state, and county where the project is located (for some
of these variables, you may use the first occurrence).
```{r}
wt1 <- WindTurbineFinal %>%
       group_by(eia_id) %>%
       summarise(mean_pcap = mean(p_cap),
            tot_pturbine = sum(p_tnum),
            mean_tcap=mean(t_cap),
            mean_thh=mean(t_hh),
            mean_trd=mean(t_rd))

wt2 <- WindTurbineFinal %>%
       select(eia_id, p_year, t_state, t_county) %>%
       distinct()

wt2 <- wt2 %>%
       distinct(eia_id, .keep_all = TRUE)

WindTP <- merge(wt1, wt2, by='eia_id')

```
# 8. Create a summary table for the data at the project level. Report the mean, standard
deviation, min, and max of mean project capacity, average number of turbines in
a project, average hub-height, average rotor diameter. Report the total number of
projects (i.e. N)?

```{r}
labs1 <- c('Mean project capacity', 'Average number of turbines in a project',
    'Average hub-height', 'Average rotor diameter' )
sumtable(WindTP, vars = c('mean_pcap','tot_pturbine', 'mean_thh', 'mean_trd'),
         summ=c('notNA(x)', 'mean(x)', 'sd(x)', 'min(x)', 'max(x)'),
         summ.names = list(c('N','Mean','SD','Min','Max')),
         labels=labs1)
```
# Remove datasets not required 
```{r}
rm(WindTurbineR, WindTurbine1, WindTurbine2, wt1, wt2 )
```
## 9. While using the dataset at the project level, create a plot(s) (using ggplot) showing
the evolution of mean hub-height and mean rotor diameter over the years (2001 -
2022). Provide a brief interpretation of your findings. *

```{r}
# Aggregate the project level data over years
WTP_Year <- WindTP %>% 
            group_by(p_year) %>%
            summarize(mean.thh=mean(mean_thh),
                      mean.trd=mean(mean_trd))
  
# Evaluation of mean hub-height over years
ggplot(data = WTP_Year) +
  geom_point(mapping = aes(x = p_year , y = mean.thh), col="blue") +
  labs(x="Years", y="Mean hub-height", title="Evaluation of mean hub-height over years") + 
  theme_bw()

# Evaluation of mean rotor diameter over years
ggplot(data = WTP_Year) +
  geom_point(mapping = aes(x = p_year , y = mean.trd), col="blue") +
  labs(x="Years", y="Mean rotor diameter", title="Evaluation of mean rotor diameter over years") + 
  theme_bw()
```
## 10. Using the cleaned turbine data and the project level data, answer the following
questions (report all of these information in a well formatted table): *
(a) Calculate the top five states with the most wind turbines.

```{r}
# Use the cleaned turbine data to calculate the total number of wind turbines at state level
WT_state <- WindTurbineFinal %>% 
            group_by(t_state) %>%
            summarize(tot_tstate=sum(p_tnum))


```
# (b) Calculate the top five states with the highest capacity of wind projects (in
MW). Show the total capacity of that state and the US for comparison.
```{r}
# Use project level data
WTP_state <-  WindTurbineFinal %>% 
            group_by(t_state) %>%
            summarize(t_pcap=sum(p_cap))

WTP_state$t_us_pcap= sum(WTP_state$t_pcap)
```
#####################################################################################

### 2 Wind Ordinance data

# Remove observations before 2001.
```{r}
WindOrdiR <- read_csv('C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Data/wind_ordinance_main.csv')

WindOrdinanceFinal<- WindOrdiR %>% filter(ordinance_year>2000)

```
## 3. Calculate the 3 states with the most number of wind ordinances and the 3 states
with least number of wind ordinances. Compare this with the total wind capacity
(MW) in these states. Is there any meaningful conclusion we can draw from this
comparison? Why or why not? *
```{r}
WindOrdinanceFinal <- WindOrdinanceFinal %>%
              add_count(State, name = 'tot_ordinance')

# Sort the data with total number of ordinance
WO_totord <- WindOrdinanceFinal %>% arrange(desc(tot_ordinance))
```
## 4. Collapse this dataset at the year level. Create a graph (using ggplot) that shows
two lines: *
(a) cumulative number of ordinances in each year, and
(b) total new ordinances enacted each year
Compare the two lines and provide a brief interpretation (2-3 sentences).

```{r}
# Collapse this dataset at the year level
wo_col_yr <- WindOrdinanceFinal %>% 
            group_by(ordinance_year) %>%
            tally()

wo_col_yr <- rename(wo_col_yr, tot_ordinance=n)

# Number of ordinances in each year
ggplot(data=wo_col_yr, aes(x=ordinance_year, y=tot_ordinance)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=tot_ordinance), vjust=1.6, color="white", size=3.5)+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  scale_x_continuous(labels = 2001:2023, breaks = 2001:2023) +
  labs(x="Ordinance Year", y="Number of ordinances", title="Number of ordinances in each year")

# Create a variable for cumulative ordinance
wo_col_yr <- wo_col_yr %>% mutate(cum_ordinace = cumsum(tot_ordinance))

# Cumulative number of ordinances in each year
ggplot(data=wo_col_yr, aes(x=ordinance_year, y=cum_ordinace)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=cum_ordinace), vjust=1.6, color="white", size=3.5)+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  scale_x_continuous(labels = 2001:2023, breaks = 2001:2023) +
  labs(x="Year", y="Cumulative umber of ordinances", title="Cumulative number of ordinances in each year")
```
#(b) total new ordinances enacted each year 
```{r}
# create a lag variable
#wo_col_yr1 <- setDT(wo_col_yr)[, c("ordi_lastyr") := .(shift(tot_ordinance, 1L, fill = NA, type = "lag"))]


           
ggplot(data=wo_col_yr, aes(x=ordinance_year, y=tot_ordinance)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=tot_ordinance), vjust=1.6, color="white", size=3.5)+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  scale_x_continuous(labels = 2001:2023, breaks = 2001:2023) +
  labs(x="Ordinance Year", y="New ordinances enacted", title="Total new ordinances enacted each year ")
```

### 3 Wind Resource Quality data

# Download the csv file wtk_site_metadata.csv from Canvas and bring it in R.
```{r}
wtk_site_metadata <- read_csv('C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Data/wtk_site_metadata.csv')
```

#3. Remove all off-shore observations (i.e. unknown state, county, or “offshore” power
curve).
4. Remove observations in Alaska and Hawaii.
```{r}
wtk_site_c1 <- wtk_site_metadata %>% filter(State!="Unknown" | County!="Unknown" | power_curve!="offshore") 

wtk_site_c1 <- wtk_site_c1 %>% filter(!State %in% c("Alaska","Hawaii"))

```
# 5. Create a summary table with mean, standard deviation, min, and max of wind
speed (m/s), capacity factor, fraction of usable area. 

```{r}
labs2 <- c('Wind speed (m/s)', 'Capacity factor', 'Fraction of usable area' ) 

sumtable(wtk_site_c1, vars = c('wind_speed','capacity_factor', 'fraction_of_usable_area'),
         summ=c('mean(x)', 'sd(x)', 'min(x)', 'max(x)'),
         summ.names = list(c('Mean','SD','Min','Max')),
         labels=labs2)
```
# 6. List the top five states with the highest wind speeds and capacity factors. Compare
this with the states with the highest wind project capacity (from question 1.10).
What do you notice?

```{r}
# Sort the data with wind speeds and capacity factors
wtk_site_c2 <- wtk_site_c1 %>% 
            group_by(State) %>%
            summarize(mean_wind_speed=mean(wind_speed),
                      mean_capacity_factor=mean(capacity_factor))

wtk_site_cwp <- wtk_site_c2 %>% arrange(desc(mean_wind_speed))

wtk_site_ccf <- wtk_site_c2 %>% arrange(desc(mean_capacity_factor))

```
### 4 Merging the datasets

```{r}
# Wind ordinance data contains only the abbreviation for the states.We need to merge the data which include name and abbreviation for the state.

State <- read_csv("C:/Users/rosha/Dropbox (GaTech)/Georgia Tech/Spring 2024/Econ 8803 ML for Econ/Problem sets/PS1/states.csv")
```
```{r}
# merge the Wind turbine data to have a column with state abbreviation
WindTurnineFinal <- left_join(WindTurbineFinal, State, by = c("t_state" = "Abbreviation"))

WindTurnineFinal1 <-WindTurnineFinal %>% mutate(County = word(t_county , 1  , -2))

```
# Merge the wind turbines and wind ordinance datasets on the basis of state and
county variables.3 Note that county names across the states may not be unique, so,
keep this in mind while writing the code for merging.
```{r}
Wind_turb_ordi <- full_join(WindTurnineFinal1, WindOrdinanceFinal,by=c("State","County"))
```
# 2. Certain observations in the merged dataset might have NAs in the ordinance_year
field. Create a binary variable that tracks which observations have an ordinance in
a specific year.
```{r}
Wind_turb_ordi$oy2001 <- ifelse(Wind_turb_ordi$ordinance_year==2001,1,0)
Wind_turb_ordi$oy2002 <- ifelse(Wind_turb_ordi$ordinance_year==2002,1,0)
Wind_turb_ordi$oy2003 <- ifelse(Wind_turb_ordi$ordinance_year==2003,1,0)
Wind_turb_ordi$oy2004 <- ifelse(Wind_turb_ordi$ordinance_year==2004,1,0)
Wind_turb_ordi$oy2005 <- ifelse(Wind_turb_ordi$ordinance_year==2005,1,0)
Wind_turb_ordi$oy2006 <- ifelse(Wind_turb_ordi$ordinance_year==2006,1,0)
Wind_turb_ordi$oy2007 <- ifelse(Wind_turb_ordi$ordinance_year==2007,1,0)
Wind_turb_ordi$oy2008 <- ifelse(Wind_turb_ordi$ordinance_year==2008,1,0)
Wind_turb_ordi$oy2009 <- ifelse(Wind_turb_ordi$ordinance_year==2009,1,0)
Wind_turb_ordi$oy2010 <- ifelse(Wind_turb_ordi$ordinance_year==2010,1,0)
Wind_turb_ordi$oy2011 <- ifelse(Wind_turb_ordi$ordinance_year==2011,1,0)
Wind_turb_ordi$oy2012 <- ifelse(Wind_turb_ordi$ordinance_year==2012,1,0)
Wind_turb_ordi$oy2013 <- ifelse(Wind_turb_ordi$ordinance_year==2013,1,0)
Wind_turb_ordi$oy2014 <- ifelse(Wind_turb_ordi$ordinance_year==2014,1,0)
Wind_turb_ordi$oy2015 <- ifelse(Wind_turb_ordi$ordinance_year==2015,1,0)
Wind_turb_ordi$oy2016 <- ifelse(Wind_turb_ordi$ordinance_year==2016,1,0)
Wind_turb_ordi$oy2017 <- ifelse(Wind_turb_ordi$ordinance_year==2017,1,0)
Wind_turb_ordi$oy2018 <- ifelse(Wind_turb_ordi$ordinance_year==2018,1,0)
Wind_turb_ordi$oy2019 <- ifelse(Wind_turb_ordi$ordinance_year==2019,1,0)
Wind_turb_ordi$oy2020 <- ifelse(Wind_turb_ordi$ordinance_year==2020,1,0)
Wind_turb_ordi$oy2021 <- ifelse(Wind_turb_ordi$ordinance_year==2021,1,0)
Wind_turb_ordi$oy2022 <- ifelse(Wind_turb_ordi$ordinance_year==2022,1,0)
Wind_turb_ordi$oy2023 <- ifelse(Wind_turb_ordi$ordinance_year==2023,1,0)
```
# 3. Collapse this data to the county level. Make sure to preserve the temporal dimensions
of the data.
```{r}
#  Collapse this data to the county level 
Wind_turb_ordi_county <- Wind_turb_ordi %>%
                   group_by(County, p_year, ordinance_year) %>%
                  summarize(mean_tcap=mean(t_cap))
```


# 4. For this dataset at the county level, create a graph that shows the average wind capacity (MW) of counties with and without a wind ordinance over the years (from2001 to 2022). What do you notice?`
```{r}
Wind_turb_ordi1 <-  subset(Wind_turb_ordi,
                     subset = !is.na(Wind_turb_ordi$ordinance_year))
                     
#  Collapse this data to the county level 
Wind_turb_ordi1 <- Wind_turb_ordi1 %>%
                   group_by(p_year) %>%
                  summarize(mean_tcap=mean(t_cap))

# First drop the observations with counties with no ordinance year
Wind_turb_ordi2 <-  subset(Wind_turb_ordi,
                     subset = is.na(Wind_turb_ordi$ordinance_year))

#  Collapse this data to the county level 
Wind_turb_ordi2 <- Wind_turb_ordi2 %>%
                   group_by(p_year) %>%
                   summarize(mean_tcap=mean(t_cap))

#  create a graph that shows the average wind capacity (MW) of counties with a wind ordinance
ggplot() + 
  geom_line(data = Wind_turb_ordi1, aes(x = p_year, y = mean_tcap, color = "With Ordinance Year")) +
  geom_line(data = Wind_turb_ordi2, aes(x = p_year, y = mean_tcap, color = "Without  Ordinance Year")) + 
  labs(color = "Legend") + 
  theme_bw() +
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) +
  scale_x_continuous(labels = 2001:2023, breaks = 2001:2023) +
  labs(x="Project Year", y="The average wind capacity (MW)")

```



# 5. Now, collapse the merged dataset in (2). at the project level. Create a summary
table that could answer the following question: *
• Are projects in a county with an ordinance bigger in size (nameplate capacity),
and use more technologically advanced turbines (i.e. bigger turbines
measured by rotor diameter)?

```{r}
WTOP_county <- Wind_turb_ordi %>% 
            group_by(eia_id, County) %>%
            summarize(mean.thh=mean(t_hh),
                      mean.trd=mean(t_rd),
                      tot.ordinance=sum(tot_ordinance),
                      tot_pcap=sum(p_cap))

# Create a subdata from above for the counties with ordinance year
WTOP_county_ord <- subset(WTOP_county,
                     subset = !is.na(WTOP_county$tot.ordinance))

# Create a subdata from above for the counties with no ordinance year
WTOP_county_no_ord <- subset(WTOP_county,
                     subset = is.na(WTOP_county$tot.ordinance))

# Sum stats for the counties with ordinance year
labs3 <- c('Average turbine hub-height', 'Average rotor diameter', 'Total project capacity') 

sumtable(WTOP_county_ord, vars = c('mean.thh','mean.trd', 'tot_pcap'),
         summ=c('mean(x)', 'median(x)', 'sd(x)', 'min(x)', 'max(x)'),
         summ.names = list(c('Mean','Median', 'SD','Min','Max')),
         labels=labs3)

```
```{r}

# Sum stats for the counties with ordinance year
sumtable(WTOP_county_no_ord, vars = c('mean.thh','mean.trd', 'tot_pcap'),
         summ=c('mean(x)', 'median(x)', 'sd(x)', 'min(x)', 'max(x)'),
         summ.names = list(c('Mean','Median', 'SD','Min','Max')),
         labels=labs3)
```


